### 2014-04-11 ApTranscriptome_TR.Rmd
Updates to manuscript

interaction.transcripts <- RxNlmAIC[which(Map(grepFunc, RxNlmAIC, term = "colonyAr:") == TRUE)]
other.transcripts <- RxNlmAIC[setdiff(names(RxNlmAIC), names(interaction.transcripts))]
temperature.transcripts <- other.transcripts[which(Map(grepFunc, other.transcripts, term = "val") == TRUE)]
colony.transcripts <- other.transcripts[setdiff(names(other.transcripts), names(temperature.transcripts))]
responsive.transcripts <- c(temperature.transcripts, interaction.transcripts)
rm(other.transcripts)
```

```{r table_results, echo=FALSE, results='asis'}
clist <- c('Total', 'Colony', 'Temperature', 'Temperature:Colony')
siglist <- c(length(RxNlmAIC),
             length(colony.transcripts),
             length(temperature.transcripts),
             length(interaction.transcripts)
             )

@@

```{r expression_shape, eval=TRUE, echo=TRUE}
# calculate response type for responsive transcripts
interaction.response.type <- ldply(interaction.transcripts, .progress = "text", .inform = TRUE, RxNtype)
stopifnot(nrow(interaction.response.type) == length(interaction.transcripts))

# calculate response types for transcripts without interactions
temperature.response.type <- ldply(temperature.transcripts, .progress = "text", .inform = TRUE, RxNtype)
stopifnot(nrow(temperature.response.type) == length(temperature.transcripts))

@@

Table 4 shows the number of transcripts that fall into each expression type for each each colony. The totals for each colony include the `r nrow(temperature.transcripts)` transcripts that have consistent temperature responses between the two colonies.

**2014-04-02 -- CURRENTLY BROKEN**

```{r shiny_file, eval=FALSE, echo=FALSE}
shiny.transcripts <- TPM.dt.sub[Ap.response.type$Transcript]

# scale expression values 
shiny.transcripts[,pTPM.scaled:=scale(pTPM), by = Transcript]
str(responsive.transcripts.TPM)
write.csv(responsive.transcripts.TPM, file = "ApRxN-shinyapp/responsive.transcripts.TPM.csv", quote = TRUE, row.names = FALSE)
```

@@

png(paste(resultsdir, "PDF_expression_all.png", sep=""))
g3 <- ggplot(Ap.df, aes(x=max.val, fill=colony)) + geom_density(alpha=0.2, position="identity")
g3 + scale_y_continuous(name="Density") +
  scale_x_continuous(name="Temperature of maximum expression")
dev.off()
```

@@

```{r offset, echo=TRUE, eval=TRUE}
varshape.out <- ldply(interaction.transcripts, .progress = "text", RxNvarshape)

@@

```{r echo=FALSE, results='hide'}
responsive.transcripts.ann.type <- signif.transcripts[names(responsive.transcripts)]
# combine responsive.transcripts.ann with "type" information
responsive.transcripts.ann.type <- merge(responsive.transcripts.ann, Ap.response.type[,c("Transcript", "A22.type", "Ar.type")], by = "Transcript", all = TRUE)
str(responsive.transcripts.ann.type)
write.csv(responsive.transcripts.ann.type, file = "results/Ap_responsive_genes_20140409.csv", row.names = FALSE)
```

@@

```{r GO_contingency_test}
# GO 'response to stress' hits in responsive transcripts
GO0006950.responsive <- responsive.transcripts.ann.type[grep("GO:0006950", responsive.transcripts.ann.type$GO.Biological.Process), list(Transcript, best.hit.to.nr, A22.type, Ar.type)]

@@

# Chi-square test to see if 'response to stress' related genes overrepresented in responsive.transcripts compared to full list
resp.stress.responsive.count <- nrow(responsive.transcripts.ann.type[grep("GO:0006950", responsive.transcripts.ann.type$GO.Biological.Process), list(Transcript, best.hit.to.nr)])
# GO 'response to stress' hits in all transcripts
resp.stress.all.count <- nrow(annotationtable[grep("GO:0006950", annotationtable$GO.Biological.Process), list(Sequence.Name, best.hit.to.nr)])

GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.transcripts.ann.type) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotationtable) - resp.stress.all.count), nrow = 2)

@@

```{r grep_shock}
hsp_responsive <- responsive.transcripts.ann.type[grep("shock", responsive.transcripts.ann.type$best.hit.to.nr), list(Transcript, best.hit.to.nr, A22.type, Ar.type)]
hsp_responsive

@@

Visualize differens in GO terms among expression types using a wordcloud.

```{gsea_wordcloud, cache=TRUE}
# create Corpus for text mining
GO.dfs <- c(paste(Ap.low.BP.ResTable$Term, collapse=" "), paste(Ap.hig.BP.ResTable$Term, collapse=" "),
            paste(Ap.int.BP.ResTable$Term, collapse=" "), paste(Ap.bim.BP.ResTable$Term, collapse=" "))

vs <- VectorSource(GO.dfs)
GO.corp <- Corpus(vs)
GO.corp

GO.corp <- tm_map(GO.corp, removePunctuation)

# create matrix of terms
term.matrix <- TermDocumentMatrix(GO.corp)
term.matrix <- as.matrix(term.matrix)
colnames(term.matrix) <- c('low', 'high', 'intermediate', 'bimodal')

# comparison cloud
comparison.cloud(term.matrix, max.words=200, random.order=FALSE, scale=c(0.25, 0.5))
```

## Comparison of expression patterns between colonies

A set of `r nrow(interaction.transcripts)` transcripts have expression patterns that depend on colony. In this section, I examine how these transcripts differ between colonies. Specifically, I expect genes that are up-regulated at high temperatures in the northern *A22* colony to be at constitutively higher expression in the more southern *Ar* colony. Conversely, I will examine the extent to which genes upregulated at colder temperatures in the *Ar* colony are constitutively expressed in the *A22* colony.

@@

Make plots for all significant transcripts

```{r plot_responsive, echo=FALSE, eval=FALSE, cache=TRUE}
# Line plot, expression against temp, faceted by colony
p1 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
  geom_line() +
  facet_grid(. ~ colony) +
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
p1

# Smooth plot, expression against temp, faceted by colony
p2 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
  geom_smooth() +
  facet_grid(. ~ colony) +
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
p2

# Same as p1, faceted by expression type
p3 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
  geom_line() +
  facet_grid(A22.exp.type ~ colony) +
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
p3

# Smooth plot, expression against temp, faceted by colony
p4 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
  geom_smooth() +
  facet_grid(A22.exp.type ~ colony) +
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
p4
```

Make plots for transcripts with significant temperature by colony interaction

```{r plot_interaction_responsive, echo=FALSE, eval=FALSE, cache=TRUE}
# Line plot, expression against temp, faceted by colony
p1 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
  geom_line() +
  facet_grid(. ~ colony) +
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
p1

# Smooth plot, expression against temp, faceted by colony
p2 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
  geom_smooth() +
  facet_grid(. ~ colony) +
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
p2

# Same as p1, faceted by expression type
p3 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
  geom_line() +
  facet_grid(A22.exp.type ~ colony) +
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
p3

# Smooth plot, expression against temp, faceted by colony
p4 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
  geom_smooth() +
  facet_grid(A22.exp.type ~ colony) +
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
p4
```

```{r plot, echo=FALSE, eval=FALSE, cache=TRUE}
# Line plot, expression against temp, faceted by colony
png(paste(resultsdir, "Ap_expression_by_colony_line.png", sep=""))
  p1 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
    geom_line() +
    facet_grid(. ~ colony) +
    scale_y_continuous(name="Expression (scaled)") +
    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
  p1
dev.off()

# Smooth plot, expression against temp, faceted by colony
png(paste(resultsdir, "Ap_expression_by_colony_smooth.png", sep=""))
  p2 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
    geom_smooth() +
    facet_grid(. ~ colony) +
    scale_y_continuous(name="Expression (scaled)") +
    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
  p2
dev.off()

# Same as p1, faceted by expression type
png(paste(resultsdir, "Ap_expression_by_colony_exp_line.png", sep=""))
  p3 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
    geom_line() +
    facet_grid(A22.exp.type ~ colony) +
    scale_y_continuous(name="Expression (scaled)") +
    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
  p3
dev.off()

# Smooth plot, expression against temp, faceted by colony
png(paste(resultsdir, "Ap_expression_by_colony_exp_smooth.png", sep=""))
  p4 <- ggplot(temperature.transcripts.TPM, aes(x=val, y=exp.scaled, group=Transcript)) +
    geom_smooth() +
    facet_grid(A22.exp.type ~ colony) +
    scale_y_continuous(name="Expression (scaled)") +
    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
  p4
dev.off()

```

```{r plot_interaction_responsive_to_file, echo=FALSE, eval=FALSE, cache=TRUE}
# Line plot, expression against temp, faceted by colony
png(paste(resultsdir, "Ap_expression_interaction_by_colony_line.png", sep=""))
  p1 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
    geom_line() +
    facet_grid(. ~ colony) +
    scale_y_continuous(name="Expression (scaled)") +
    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
  p1
dev.off()

# Smooth plot, expression against temp, faceted by colony
png(paste(resultsdir, "Ap_expression_interaction_by_colony_smooth.png", sep=""))
  p2 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
    geom_smooth() +
    facet_grid(. ~ colony) +
    scale_y_continuous(name="Expression (scaled)") +
    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
  p2
dev.off()

# Same as p1, faceted by expression type
png(paste(resultsdir, "Ap_expression_interaction_by_colony_exp_line.png", sep=""))
  p3 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
    geom_line() +
    facet_grid(A22.exp.type ~ colony) +
    scale_y_continuous(name="Expression (scaled)") +
    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
  p3
dev.off()

# Smooth plot, expression against temp, faceted by colony
png(paste(resultsdir, "Ap_expression_interaction_by_colony_exp_smooth.png", sep=""))
  p4 <- ggplot(temperature.transcripts.TPM.interaction, aes(x=val, y=exp.scaled, group=Transcript)) +
    geom_smooth() +
    facet_grid(A22.exp.type ~ colony) +
    scale_y_continuous(name="Expression (scaled)") +
    scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
  p4
dev.off()
```

>>>>

interaction.lms <- RxNlmAIC[which(Map(grepFunc, RxNlmAIC, term = "colonyAr:") == TRUE)]
other.lms <- RxNlmAIC[setdiff(names(RxNlmAIC), names(interaction.lms))]
temperature.lms <- other.lms[which(Map(grepFunc, other.lms, term = "val") == TRUE)]
colony.lms <- other.lms[setdiff(names(other.lms), names(temperature.lms))]
responsive.lms <- c(temperature.lms, interaction.lms)
rm(other.lms)
```

```{r table_results, echo=FALSE, results='asis'}
clist <- c('Total', 'Colony', 'Temperature', 'Temperature:Colony')
siglist <- c(length(RxNlmAIC),
             length(colony.lms),
             length(temperature.lms),
             length(interaction.lms)
             )

@@

```{r expression_shape, eval=TRUE, echo=TRUE}
# calculate response type for responsive transcripts
interaction.response.type <- ldply(interaction.lms, .progress = "text", .inform = TRUE, RxNtype)
stopifnot(nrow(interaction.response.type) == length(interaction.lms))

# calculate response types for transcripts without interactions
temperature.response.type <- ldply(temperature.lms, .progress = "text", .inform = TRUE, RxNtype)
stopifnot(nrow(temperature.response.type) == length(temperature.lms))

@@

Table 4 shows the number of transcripts that fall into each expression type for each each colony. The totals for each colony include the `r nrow(temperature.lms)` transcripts that have consistent temperature responses between the two colonies.

@@

png(paste(resultsdir, "PDF_expression_all.png", sep=""))
g3 <- ggplot(Ap.df, aes(x=max.val, fill=colony)) + 
  geom_density(alpha=0.2, position="identity") + 
  #scale_fill_manual(name = "Colony", values = c("white", "black")) + 
  scale_y_continuous(name="Density") +
  scale_x_continuous(name="Temperature of maximum expression")
suppressWarnings(print(g3))
dev.off()
```

@@

```{r offset, echo=TRUE, eval=TRUE}
varshape.out <- ldply(interaction.lms, .progress = "text", RxNvarshape)

@@

```{r echo=FALSE, results='hide'}
responsive.lms.ann.type <- signif.transcripts[names(responsive.lms)]
# combine responsive.lms.ann with "type" information
responsive.lms.ann.type <- merge(responsive.lms.ann.type, Ap.response.type[,c("Transcript", "A22.type", "Ar.type")], by = "Transcript", all = TRUE)
str(responsive.lms.ann.type)
write.csv(responsive.lms.ann.type, file = "results/Ap_responsive_genes_20140409.csv", row.names = FALSE)
```

@@

```{r GO_contingency_test}
# GO 'response to stress' hits in responsive transcripts
GO0006950.responsive <- responsive.lms.ann.type[grep("GO:0006950", responsive.lms.ann.type$GO.Biological.Process), list(Transcript, best.hit.to.nr, A22.type, Ar.type)]

@@

# Chi-square test to see if 'response to stress' related genes overrepresented in responsive.lms compared to full list
resp.stress.responsive.count <- nrow(responsive.lms.ann.type[grep("GO:0006950", responsive.lms.ann.type$GO.Biological.Process), list(Transcript, best.hit.to.nr)])
# GO 'response to stress' hits in all transcripts
resp.stress.all.count <- nrow(annotationtable[grep("GO:0006950", annotationtable$GO.Biological.Process), list(Sequence.Name, best.hit.to.nr)])

GO.stress.table <- matrix(rbind(resp.stress.responsive.count, nrow(responsive.lms.ann.type) - resp.stress.responsive.count, resp.stress.all.count, nrow(annotationtable) - resp.stress.all.count), nrow = 2)

@@

```{r grep_shock}
hsp_responsive <- responsive.lms.ann.type[grep("shock", responsive.lms.ann.type$best.hit.to.nr), list(Transcript, best.hit.to.nr, A22.type, Ar.type)]
hsp_responsive

@@

@@

Organize data

```{r plot_data}
# extract TPM data for thermally-responsive transcripts
resp.TPM.dt.sub <- TPM.dt.sub[names(responsive.lms)]
setkey(resp.TPM.dt.sub, Transcript)
str(resp.TPM.dt.sub)
length(unique(resp.TPM.dt.sub$Transcript))
# scale transcripts so can compare
resp.TPM.dt.sub[,TPM.scaled:=scale(TPM), by = Transcript]
```

Make plots for all genes expressed at *High* temps in GO category "GO:0006950: response to stress"

```{r plot_GOstress, echo=FALSE, eval=FALSE, cache=TRUE}
high.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("High", Ar.type)), with(GO0006950.responsive, grep("High", A22.type))), ]

# Line plot, expression against temp, faceted by colony
trp1 <- ggplot(resp.TPM.dt.sub[high.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
  geom_line() + 
  facet_grid(. ~ colony) + 
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
print(trp1)

# Smooth plot, expression against temp, faceted by colony
trp2 <- ggplot(resp.TPM.dt.sub[high.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
  facet_grid(. ~ colony) + 
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
print(trp2)
```

Make plots for all genes expressed at *Low* temps in GO category "GO:0006950: response to stress"

```{r plot_GOstress, echo=FALSE, eval=FALSE, cache=TRUE}
low.stress.terms <- GO0006950.responsive[intersect(with(GO0006950.responsive, grep("Low", Ar.type)), with(GO0006950.responsive, grep("Low", A22.type))), ]

# Line plot, expression against temp, faceted by colony
trp3 <- ggplot(resp.TPM.dt.sub[low.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
  geom_line() + 
  facet_grid(. ~ colony) + 
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
print(trp3)

# Smooth plot, expression against temp, faceted by colony
trp4 <- ggplot(resp.TPM.dt.sub[low.stress.terms$Transcript], aes(x=val, y=TPM.scaled, group=Transcript)) + 
  geom_smooth(method = "lm", formula = y ~ poly(x, 2)) + 
  facet_grid(. ~ colony) + 
  scale_y_continuous(name="Expression (scaled)") +
  scale_x_continuous(name=expression(paste("Temperature ", degree, "C")))
print(trp4)
```

```{r shiny_file, eval=FALSE, echo=FALSE}
shiny.dt <- TPM.dt.sub[Ap.response.type$Transcript]

# scale expression values 
shiny.dt[,TPM.scaled:=scale(TPM), by = Transcript]
str(shiny.dt)
write.csv(shiny.dt, file = "ApRxN-shinyapp/responsive.lms.TPM.csv", quote = TRUE, row.names = FALSE)
```

### 2014-04-11 doc/ApTranscriptome_manuscript.Rmd
Updates to manuscript

The study of whole organism gene expression (e.g. transcriptomics) has rapidly increased due to improved sequencing technology (RNAseq). To date, RNAseq studies have largely focused on differential expression across few environments [@teets2012, @pespeni2013, @fu2014] or tissues [@li2013] using an ANOVA-type statistical approach. However, many ecologically-relevant aspects of the environment such as temperature vary continuously in nature. Variation in the phenotype of an organism in response to such a continuous predictor (i.e. phenotypic plasticity) is referred to as a reaction norm. A reaction norm can differ not only in mean values, but also in the shape of the curve as is often found for thermal reaction norms [@hoffmann2013; @murren2014]. An analysis of phenotypic reaction norms found that differences in the shape were greater than differences in mean values in comparisons among species and populations [@murren2014]. The authors concluded that many patterns of biologically-relevant plasticity are likely to be missed in studies with few environments. Similarly, gene expression studies with few environments are likely to miss many biologically-relevant differences in gene expression. To overcome this limitation, we use a novel method to characterize the reaction norm of all genes in the transcriptome (e.g. the reactionome) in response to thermal stress. Specifically, we measure gene expression across a thermal gradient and identify thermally-responsive genes using a regression approach.

We apply this approach in a study of ants in the *Aphaenogaster rudis-picea* species complex [@umphrey1996]. Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to litter processing and decomposition [@lyford1963; @folgarait1998, @terbough2001, @holec2005]. As ectothermic species, ants experience a wide range of temperature conditions. Throughout the year, ant colonies persist through freezings and hot summers, and at shorter time scales, individual workers can experience substantial thermal shock as they leave the buffered ant nest to forage for food [...]. Thus, at the organismal level we expect that ants utilize multiple molecular pathways to deal tolerate thermal stress.

To explore the molecular toolkit used by *Aphaenogaster* to cope with thermal stress, we performed a whole organism RNA sequencing study using environmentally-divergent colonies from North Carolina and Vermont, USA. Our specific goals were to 1) generate the set of all genes (i.e. transcriptome) expressed by *Aphaenogaster* across physiologically-relevant temperatures, 2) identify genes that have responsive expression across this thermal gradient (i.e. thermal reactionome), 3) explore the functional annotations of genes that are activated or down-regulated in response to cold and high temperatures, and 4) examine differences in the thermal reactionome between the two colonies.

@@

to each gene. We used $log(TPM + 1)$ as the response to control for skew in the expression data. For a continous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects even [@cottingham2005]. This method is robust to overdispersion as we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature. To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining genes under the 5% FDR threshold.

We identified thermally-responsive genes as the subset under our FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after stepwide model selection by AIC. All analyses were performed in R [@RCoreTeam2014] and are fully reproducible (Supplemental file).

@@

To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two colonies using the program `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossman2007] implemented in R [@RCoreTeam2014]. Briefly, this approach identifies GO terms that are overrepresented in the significant genes relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.

@@

```{r load, echo=FALSE, results='hide'}
load("../ms.RData")
suppressWarnings(library(pander))
suppressWarnings(library(ggplot2))
```

@@

```{r expression_type_table, echo=FALSE, results='asis'}
pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each colony and their overlap.")
```

@@

We identified 22,495 genes with overall model fits of P < 0.05, retaining 9,809 after correcting for FDR. Of these transripts, 8,044 significantly change expression with temperature including 6,727 that have a colony x temperature interaction. The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not expressed differed significantly between the colonies ($\chi^2 = 170, P < 0.0001$; Table 1). For both colonies, the majority of responsive transcripts were in the *Low* category (Fig. 1). More transcripts had *Intermediate* expression in *ApNC* than in *ApVT*, which had nearly twice as many bimodally-expressed transcripts  (Fig. 1, Table 1).

@@

We quantified differences in the reaction norms of thermally-responsive genes between colonies as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta W$) and wiggle ($\Delta W$). Overall, we found that there was no significant difference in position of the mean (*t* = -0.95, *P* = 0.34), but changes in the shape due to slope (*t* = 24.67, *P* < 0.001), curvature (*t* = 15.71, *P* < 0.001) and wiggle (*t* = -12.45, *P* < 0.001) were all highly signifcant. On average, slope (95% CI 0.41 - 0.48) and curvature (95% CI 0.25 - 0.33) were greater in *ApVT* than *ApNC*, indicating a steeper response to temperature.

@@

Plasticity is a well-characterized mechanism by which organims tolerate environmental change. Given our sampling of a warmer (*ApNC*) and colder (*ApVT*) climate species, we hypothesized that genes which are plastically-expressed under heat shock in the colder species would have greater constitutive expression in the warmer species, and vice versa. Consistent with one of our predictions, we found that constitutive expression of *High* genes from *ApVT* was greater in the more southern *ApNC* colony (Wilcoxon V = 911970, *P* < 0.001, 95% CI location shift -0.11 - 0.03). However, for genes with *Low* expression in *ApNC* we also found that constitutive expression was greater in *ApNC* than *ApVT* (Wilcoxon V = 2041149, *P* < 0.001, 95% CI location shift -0.23 - 0.15).

@@

Previous studies have using *Drosophila* [@sorensen2005] and a temperate flesh fly [@teets2012b] have implicated GO terms .... as responding to heat stress. In this study, we found .... of these GO terms to be enriched at high temperatures for *Aphaenogaster*. However, when we looked at the exact identify of the genes, we only found X overlapping genes.

@@

A key outcome of this study was developing the first genomic tools for the study of adaptive differentiation in *Aphaenogaster*. Our comparative results are limited to two populations, and an important future step will be expanding this work to eaexamine differences in gene expression among a larger number of populations. Moreover, the potential for species to adapt to climate change will depend in part on their genetic variation for thermal tolerance. The quantitative genetic studies required to estimate this adaptive potential have been out of reach of ant biologists due to the long lifespan of colonies, and inability of ants to mate in the lab. However, future work could leverage genomic information to infer the relatedness among colonies in the absence of a known pedigree, enabling estimation of genetic variation within populations and their adaptive potential.

>>>>

The study of whole organism gene expression (e.g. transcriptomics) has rapidly increased due to improved sequencing technology (RNAseq). To date, RNAseq studies have largely focused on differential expression across few environments [@teets2012; @pespeni2013; @barshis2013; @fu2014] or tissues [@li2013] using an ANOVA-type statistical approach. However, many ecologically-relevant aspects of the environment such as temperature vary continuously in nature. Variation in the phenotype of an organism in response to such a continuous predictor (i.e. phenotypic plasticity) is referred to as a reaction norm. A reaction norm can differ not only in mean values, but also in the shape of the curve as is often found for thermal reaction norms [@hoffmann2013; @murren2014]. An analysis of phenotypic reaction norms found that differences in the shape were greater than differences in mean values in comparisons among species and populations [@murren2014]. The authors concluded that many patterns of biologically-relevant plasticity are likely to be missed in studies with few environments. Similarly, gene expression studies with few environments are likely to miss many biologically-relevant changes in gene expression. To overcome this limitation, we use a novel method to characterize the reaction norm of all genes in the transcriptome (e.g. the reactionome) in response to thermal stress. Specifically, we measure gene expression across a thermal gradient and identify thermally-responsive genes using a regression approach.

We apply this approach in a study of two closely related eastern North American ant species, *Aphaenogaster rudis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively. Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguezcabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006]. As ectothermic species, ants experience a wide range of temperature conditions. Throughout the year, ant colonies persist through freezings and hot summers, and at shorter time scales, individual workers can experience substantial thermal shock as they leave the buffered ant nest to forage for food. Thus, at the organismal level we expect that ants utilize multiple molecular pathways to tolerate thermal stress.

To explore the molecular toolkit used by *Aphaenogaster* to cope with thermal stress, we performed a whole organism RNA sequencing study. Our specific goals were to 1) generate the set of all genes (i.e. transcriptome) expressed by *Aphaenogaster* across physiologically-relevant temperatures, 2) identify genes that have responsive expression across this thermal gradient (i.e. thermal reactionome), 3) explore the functional annotations of genes that are activated or down-regulated in response to cold and high temperatures. We expected to observe candidate genes in the *response to stress* Gene Ontology (GO) category including *heat shock proteins* and ... in the thermally-responsive group.

While characterizing the reactionome is useful for description of the molecular pathways for thermal tolerance, a greater understanding of the process of evolution can be gained using a comparative approach [@harvey1991]. Specifically, comparisons of populations or species in contrasting environments can reveal the extent to which geographic differentiation and selection has changed the number of genes involved in thermal tolerance. To accomplish this, we compared the thermal reactionomes between the two species of *Aphaenogaster*. *A. picea*, the northern species, experiences less chronic and acute heat stress than *A. rudis*. Thus, we that *A. picea* would have more genes than *A. rudis* that are up-regulated to high temperatures, while *A. rudis* would have higher constitutive expression of these thermally-tolerant genes. Reciprocally, we expected *A. rudis* to have greater up-regulation of genes in response to cold temperatures, while *A. picea* would have greater constitutive expression of cold-tolerant genes. Finally, phenotypic studies in *Aphaenogaster* [@warren2013] and other ecotherms [@hoffmann2013] have found less variation in critical maximum than minimum temperatures. From this, we hypothesized that if the species had more similar maximum temperatures than minimum temperatures, the northern species *A. picea* would have a broader thermal breadth of gene expression than *A. rudis*.

@@

to each gene. We used $log(TPM + 1)$ as the response to control for skew in the expression data. For a continuous predictor such as temperature, this regression approach is preferred to an ANOVA approach as it can reveal non-linear responses such as hump-shaped or threshold effects even [@cottingham2005]. This method is robust to over-dispersion as we expect errors in the read count distribution [@anders2010] to be independent with respect to temperature. To correct for multiple testing, we calculated adjusted *P* values using the False Discovery Rate (FDR) approach of Benjamini and Hochberg [-@benjamini1995], retaining genes under the 5% FDR threshold.

We identified thermally-responsive genes as the subset under our FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC. All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental file).

@@

To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two colonies using the program `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossmann2007] implemented in R [@r_core_team2013]. Briefly, this approach identifies GO terms that are overrepresented in the significant genes relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.

@@

```{r load, echo=FALSE, results='hide'}
options(scipen = 1, digits = 3, warn = -1)
load("../ms.RData")
suppressMessages(library(pander))
suppressMessages(library(ggplot2))
```

@@

```{r expression_type_table, echo=FALSE, results='asis'}
pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each colony and their overlap.")
library(xtable)
xtable(Ap.type.table)
```

@@

We identified 22,495 genes with overall model fits of P < 0.05, retaining 9,809 after correcting for FDR. Of these transcripts, 8,044 significantly change expression with temperature including 6,727 that have a colony x temperature interaction. The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not expressed differed significantly between the colonies ($\chi^2 = 170, P < 0.0001$; Table 1). For both colonies, the majority of responsive transcripts were in the *Low* category (Fig. 1). More transcripts had *Intermediate* expression in *ApNC* than in *ApVT*, which had nearly twice as many bimodally-expressed transcripts  (Fig. 1, Table 1).

@@

We quantified differences in the reaction norms of thermally-responsive genes between colonies as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta W$) and wiggle ($\Delta W$). Overall, we found that there was no significant difference in position of the mean (*t* = -0.95, *P* = 0.34), but changes in the shape due to slope (*t* = 24.67, *P* < 0.001), curvature (*t* = 15.71, *P* < 0.001) and wiggle (*t* = -12.45, *P* < 0.001) were all highly significant. On average, slope (95% CI 0.41 - 0.48) and curvature (95% CI 0.25 - 0.33) were greater in *ApVT* than *ApNC*, indicating a steeper response to temperature.

@@

Plasticity is a well-characterized mechanism by which organism tolerate environmental change. Given our sampling of a warmer (*ApNC*) and colder (*ApVT*) climate species, we hypothesized that genes which are plastically-expressed under heat shock in the colder species would have greater constitutive expression in the warmer species, and vice versa. Consistent with one of our predictions, we found that constitutive expression of *High* genes from *ApVT* was greater in the more southern *ApNC* colony (Wilcoxon V = 911970, *P* < 0.001, 95% CI location shift -0.11 - 0.03). However, for genes with *Low* expression in *ApNC* we also found that constitutive expression was greater in *ApNC* than *ApVT* (Wilcoxon V = 2041149, *P* < 0.001, 95% CI location shift -0.23 - 0.15).

@@

Previous studies have using *Drosophila* [@sorensen2005] and a temperate flesh fly [@teets2012-1] have implicated GO terms .... as responding to heat stress. In this study, we found .... of these GO terms to be enriched at high temperatures for *Aphaenogaster*. However, when we looked at the exact identify of the genes, we only found X overlapping genes.

@@

A key outcome of this study was developing the first genomic tools for the study of adaptive differentiation in *Aphaenogaster*. Our comparative results are limited to two populations, and an important future step will be expanding this work to examine differences in gene expression among a larger number of populations. Moreover, the potential for species to adapt to climate change will depend in part on their genetic variation for thermal tolerance. The quantitative genetic studies required to estimate this adaptive potential have been out of reach of ant biologists due to the long lifespan of colonies, and inability of ants to mate in the lab. However, future work could leverage genomic information to infer the relatedness among colonies in the absence of a known pedigree, enabling estimation of genetic variation within populations and their adaptive potential.
